Type Error Telemetry at Scale

### 2023-09-29 programming23/paper.tex
sec 4,5,7: creators -> sessions

Fortunately, the data rarely contains too-complex errors.
None of these errors appear in edit ranges~(\cref{t:type-error-count}).
For the specific case of \code{CodeTooComplex}, our telemetry tracked project-wide
counts and found only 26 of these errors, which were spread across eleven
records in three sessions.
It is entirely possible that all these errors came from one creator who
deliberately pushed the limits of type analysis.
Whatever the reason, the overall number is low enough to suggest that the
internal limits are not posing barriers to typechecker adoption.

@@

The apparent visual symmetry in the density plots~(\cref{f:error-density})
suggests that edits add and remove errors with rougly equal frequency,
regardless of mode.
No matter whether a creator is using \mnocheck{}, \mnonstrict{},
or \mstrict{} mode, edits seem to balance increases and decreases in the number
of type errors.
\Cref{f:error-changes} explores this question of balance in detail.
For each of the three modes, it reports the percent of edits that increase (Add),
maintain at a nonzero level (Keep), or decrease (Drop) the number of type
errors in the project.

@@

\begin{itemize}
  \item
    The Adds and Drops are remarkably balanced: across all three modes, they are
    at most \pct{2} apart from one another.
    This confirms that changes to the number of errors are balanced.
    Creators tend to fix errors highlighted by type analysis
    even in the stricter analysis modes.

@@

\todo{No identity $\Rightarrow$ no ``creators'', only sessions.}
With the data in hand, we can answer the research questions
from~\cref{s:introduction}.
The answers confirm several aspects of Luau's design, but
at the same time offer several directions for future improvements.

Luau clearly\todo{How can we say?} has a powerful typechecker that can efficiently
analyze thousands of lines of code.
But, few creators use type analysis directly and the data presents little
evidence that doing so would improve their quality of life.
We conjecture that Luau needs further tailoring to match
common idioms in Roblox scripts.
Discovering these idioms is a question for future work~(\cref{s:conclusion}).

\begin{description}
  \item[How many creators choose to use type analysis?]
    A mere \pct{10} of creators opt-in to type checking.
    Most of these use \mnonstrict{} mode, and fewer than \pct{1}
    use \mstrict{} mode.

  \item[How often do projects contain modules with different analysis modes?]
    Projects rarely combine analysis modes.
    Only 512 of \nK{340} sessions ($<\pct{0.15}$) switched to modules with different
    analysis modes.
    This number may be a significant underapproximation because we have data
    only on modules that the creator chose to open, but the widespread use
    of \mnocheck{} mode suggests that it is not off by much.

  \item[How often do creators turn analysis off?]
    Creators rarely disable type analysis after opting in.
    Only 233 of the \nK{340} sessions ($<\pct{0.07}$)
    switch to a weaker mode.
    Most of these (\pct{75}) downgrade from \mstrict{} to
    \mnonstrict{} % BEN: confirm!
    rather than from \mnonstrict{} mode.
    %% FILL more insight if we match upgrades + downgrades in session?

@@

      \item[- How do creators respond? Which errors persist through edits?]
        Creators typically fix analysis errors.
        With few exceptions, increases and decreases in the number of errors
        balance out~(\cref{f:error-changes,t:type-error-survival,}).
        Even in \mnocheck{}, most edits that overlap with syntax errors
        remove at least one error~(\pct{67}) rather than increase or maintain
        the total.
        The errors that most often survive edits are due to
        inference at a binary operation,
        sending \code{self} to a function that does not expect it,
        and failure to unpack an optional value.
        But the first two seldom arise, and the third is something creators
        might reasonably ignore while prototyping.

@@

  \item[What impact does type analysis have on the number of \FS{} errors?]
    Opting in to type analysis has little impact on \FS{} errors.
    The percent of \FS{} errors in \mnocheck{} code matches the percent of
    creators using \mnocheck{} mode~(\cref{f:error-by-mode}); whereas,
    if type analysis helped reduce \FS{} errors, \mnocheck{} would own a
    higher percentage of the errors.
    Furthermore, there is no apparent relation between \FS{} errors and edits,
    regardless of analysis mode~(\cref{f:error-changes}).
    A randomly-chosen edit has a 1/3 chance of increasing, decreasing, or
    maintaining the number of \FS{} errors.

@@

Fourth, type analysis reports several errors at once and we have no idea which
errors, if any, creators intended to fix with their edits.
Thus, fixes may be a side effect of other plans, as seen in prior work [CITE].
% Cite the Marceau SIGCSE 2011 paper. This is the [DIFF] code from the rubric.

@@

Lastly, we have two remarks about external validity.
First, our information on type errors is limited to error names.
If another language has an error that is roughly similar to
the\todo{Huh? It's similar yet different?}
\code{CannotCallNonFunction} error but with slightly different
semantics, then the relative frequency of that error could be different
than what we observed.
Second, we know nothing about the creators in our study except that they
were selected at random from a large and diverse group.
A dataset based on a targeted subset of users might show entirely different
characteristics.

@@

This paper presents the first large-scale analysis of developers' interactions with
an industrial strength typechecker.
With over 1.5 million telemetry records from over 300,000 sessions
that took place during a three-month timespan, the data has plenty to say about
type errors and their overlap with program edits.
At the same time, the data is carefully restricted to avoid leaking personal
and proprietary information: we know the number and kind of type errors
that occurred, but nothing about the line(s) of code that the typechecker
complained about.

On the object level, the analysis offers several lessons for the Luau
developers and other maintainers of gradually-typed languages:
\begin{itemize}
  \item
    Nearly all creators (\pct{90}) use the default ``no check''
    typechecking mode~(\cref{f:dataset-overview}).
    Since the next typing mode (\mnonstrict{}) requires no annotations and is
    very conservative about the errors it reports, we conjecture that
    developers would have no objections to using it and are simply unaware that
    other modes exist.
%  \item
%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    Creators tend to fix errors immediately rather than dealing with them in a batch
    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
    % Changes to the number of errors balance one another~(\cref{f:error-density});
    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
  \item
    All Luau scripts interact with data assets, but no typechecking mode
    has a precise view of the data.
    Improving precision is an important direction for future work.
    Until then, \mstrict{} mode clearly needs to relax its data model
    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
\end{itemize}

>>>>

Fortunately, the data rarely contains too-complex errors.
None of these errors appear in edit ranges~(\cref{t:type-error-count}).
For the specific case of \code{CodeTooComplex}, our telemetry tracked project-wide
counts and found only 26 of these errors, which were spread across eleven
records in three sessions.
It is entirely possible that all these errors came from one session that
deliberately pushed the limits of type analysis.
Whatever the reason, the overall number is low enough to suggest that the
internal limits are not posing barriers to typechecker adoption.

@@

The apparent visual symmetry in the density plots~(\cref{f:error-density})
suggests that edits add and remove errors with rougly equal frequency,
regardless of mode.
No matter whether a session is using \mnocheck{}, \mnonstrict{},
or \mstrict{} mode, edits seem to balance increases and decreases in the number
of type errors.
\Cref{f:error-changes} explores this question of balance in detail.
For each of the three modes, it reports the percent of edits that increase (Add),
maintain at a nonzero level (Keep), or decrease (Drop) the number of type
errors in the project.

@@

\begin{itemize}
  \item
    The Adds and Drops are remarkably balanced: across all three modes, they are
    at most \pct{2} apart from one another.
    This confirms that changes to the number of errors are balanced.
    Sessions tend to fix errors highlighted by type analysis
    even in the stricter analysis modes.

@@

With the data in hand, we can answer the research questions
from~\cref{s:introduction}.
The answers confirm several aspects of Luau's design, but
at the same time offer several directions for future improvements.

Luau clearly\todo{How can we say?} has a powerful typechecker that can efficiently
analyze thousands of lines of code.
But, few sessions use type analysis directly and the data presents little
evidence that doing so would improve creators' quality of life.
We conjecture that Luau needs further tailoring to match
common idioms in Roblox scripts.
Discovering these idioms is a question for future work~(\cref{s:conclusion}).

\begin{description}
  \item[How many creators choose to use type analysis?]
    A mere \pct{10} of sessions (and of telemetry records overall) opt-in to
    type checking~(\cref{f:dataset-overview}).
    Most of these use \mnonstrict{} mode, and fewer than \pct{1}
    use \mstrict{} mode.

  \item[How often do projects contain modules with different analysis modes?]
    Sessions rarely combine analysis modes.
    Only 512 of \nK{340} sessions ($<\pct{0.15}$) switched to modules with different
    analysis modes.
    This number may be a significant underapproximation because we have data
    only on modules that the creator chose to open, but the widespread use
    of \mnocheck{} mode suggests that it is not off by much.

  \item[How often do creators turn analysis off?]
    Sessions rarely disable type analysis after opting in.
    Only 233 of the \nK{340} sessions ($<\pct{0.07}$)
    switch to a weaker mode.
    Most of these (\pct{75}) downgrade from \mstrict{} to
    \mnonstrict{} % BEN: confirm!
    rather than from \mnonstrict{} mode.
    %% FILL more insight if we match upgrades + downgrades in session?

@@

      \item[- How do creators respond? Which errors persist through edits?]
        Sessions typically fix analysis errors.
        With few exceptions, increases and decreases in the number of errors
        balance out~(\cref{f:error-changes,t:type-error-survival,}).
        Even in \mnocheck{}, most edits that overlap with syntax errors
        remove at least one error~(\pct{67}) rather than increase or maintain
        the total.
        The errors that most often survive edits are due to
        inference at a binary operation,
        sending \code{self} to a function that does not expect it,
        and failure to unpack an optional value.
        But the first two seldom arise, and the third is something creators
        might reasonably ignore while prototyping.

@@

  \item[What impact does type analysis have on the number of \FS{} errors?]
    Opting in to type analysis has little impact on \FS{} errors.
    The percent of \FS{} errors in \mnocheck{} code matches the percent of
    sessions using \mnocheck{} mode~(\cref{f:error-by-mode}); whereas,
    if type analysis helped reduce \FS{} errors, \mnocheck{} would own a
    higher percentage of the errors.
    Furthermore, there is no apparent relation between \FS{} errors and edits,
    regardless of analysis mode~(\cref{f:error-changes}).
    A randomly-chosen edit has a 1/3 chance of increasing, decreasing, or
    maintaining the number of \FS{} errors.

@@

Fourth, type analysis reports several errors at once and we have no idea which
errors, if any, creators intended to fix with their edits.
Thus, fixes may be a side effect of other plans, as seen in prior work~\cite{mfk-sigcse-2011}.
% This is the [DIFF] code from Marceau SIGCSE 2011.

@@

Lastly, we have two remarks about external validity.
First, our information on type errors is limited to error names.
If another language has an error that is roughly similar to
the\todo{Huh? It's similar yet different?}
\code{CannotCallNonFunction} error but with slightly different
semantics, then the relative frequency of that error could be different
than what we observed.
%% ben: not sure what to say ... may remove the error names comment
Second, we know nothing about the creators in our study except that they
were selected at random from a large and diverse group.
A dataset based on a targeted subset of users might show entirely different
characteristics.

@@

This paper presents the first large-scale analysis of developers' interactions with
an industrial strength typechecker.
With over 1.5 million telemetry records from over 340,000 sessions
that took place during a three-month timespan, the data has plenty to say about
type errors and their overlap with program edits.
At the same time, the data is limited to avoid leaking personal
and proprietary information: we know the number and kind of type errors
that occurred, but nothing about the line(s) of code that the typechecker
complained about.

On the object level, the analysis offers several lessons for the Luau
developers and other maintainers of gradually-typed languages:
\begin{itemize}
  \item
    Nearly all sessions (\pct{90}) use the default ``no check''
    typechecking mode~(\cref{f:dataset-overview}).
    Since the next typing mode (\mnonstrict{}) requires no annotations and is
    very conservative about the errors it reports, we conjecture that
    developers would have no objections to using it and are simply unaware that
    other modes exist.
%  \item
%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    Sessions tend to fix errors immediately rather than dealing with them in a batch
    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
    % Changes to the number of errors balance one another~(\cref{f:error-density});
    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
  \item
    All Luau scripts interact with data assets, but no typechecking mode
    has a precise view of the data.
    Improving precision is an important direction for future work.
    Until then, \mstrict{} mode clearly needs to relax its data model
    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
\end{itemize}
